Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-18 latex/main.tex
Updates from Overleaf

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases, which can be caused by
systematic errors in genome sequences
associated with sequencing protocols \cite{Turakhia2020},
multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
alignment errors \cite{Hunt2024}).

@@

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively removed artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.

Different types of recombinants successfully removed are depicted
by the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}:
(1) Q1, 686;
(2) Q3, MNS at 13423-13424 \cite{DeMaio2024};
(3) Q4, 22194 (possibly MAFFT alignment error \cite{Hunt2024}).

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
Q2 & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
Q3 & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
Q4 & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
\end{tabularx}
\caption{
Examples of copying patterns.
Q1 to Q4 are examples drawn from
the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
Q2 illustrates a marginally good copying pattern.
Q1, Q3, and Q4 illustrate copying patterns that fail the QC thresholds.
}
\label{fig:copying_pattern_examples}
\end{figure}

>>>>

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases, which can be caused by
bioinformatics errors in genome assembly \cite{Turakhia2020},
multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
alignment errors \cite{Hunt2024}).

@@

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively removed artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.

Furthermore,
the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
depict different types of artefacts that were successfully removed.
Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
The vast majority of the QC-failing recombinants has this deletion, and
many cases are affected by a known amplicon dropout caused by this particular deletion.
Q3 shows an example artefact caused by a combination of an MNS and a deletion.

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
\end{tabularx}
\caption{
Examples of copying patterns.
Q1 to Q4 are examples drawn from
the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
Q2 illustrates a marginally good copying pattern.
Q1, Q3, and Q4 illustrate copying patterns that fail QC.
}
\label{fig:copying_pattern_examples}
\end{figure}
